\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-05-25 jss/cubble-jss.tex
section 2 tweaked

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and may have multiple variables measured at each spatial and temporal value. The analysis of spatio-temporal data can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. For these tasks to be carried out easily, there is a need for a spatio-temporal data representation that underpins various operations.

The \emph{Representing data} section of the SpatioTemproal CRAN task view \citep{ctvspatiotemporal} lists existing spatio-temporal data representation. Among them, \citet{spacetime} proposes four layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} \citep{spacetime} package with the underlying spatial and temporal class: \pkg{sp} \citep{sp} and \pkg{xts} \citep{xts}. The \pkg{stars} \citep{stars} package has been a recent implementation to handle both raster and vector spatio-temporal data using array. It depends on the newer spatial package \pkg{sf} \citep{sf} and interfaces with the widely-used tidyverse verbs.

{[}till here{]} One issue with wrangling spatio-temporal data is that the raw data does not naturally structured in any one of the representation above. Also, in data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble. For example, map data records the shape of a collection of area of interest is provided by official government website; geo-scientific data records the longitude and latitude coordinate of locations in the area along with other meta data relates to the location; and temporal variables indexed by both location and time. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data \citep{tidydata} should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} and \pkg{ggplot2} for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period. This poses the question on how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble}, to {[}overcome the problem{]} with organising variables in different levels when working with spatio-temporal data. The data structure focuses on connecting to the upstream data collection and is suitable for temporal data collected at the fixed location over time. Among the four spacetime layouts in \citet{spacetime}, it can be full grid layout, sparse grid layout, or irregular layout, but trajectory is not within the scope of this work. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With this new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section \ref{cube} discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.

@@

Spatio-temporal data can be conceptualised using a cubical data model with reference to variables, time and space. This naturally motivates the usage of array to represent and store spatio-temporal data, as evident by satellite imageries, large climate models, among others. The work by \citet{lu_multidimensional_2018} has discussed the suitability of this representation through array algebra and practical spatio-temporal tasks. \textcolor{magenta}{SZ: is this too direct?}

The cubical conceptual framework provides a generalisation to the operation can be done on the data for visualisation. A paper by \citet{bach_review_2014} has reviewed the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data'', but referring to ``an abstract 2D substrate that is used to visualize data at a specific time''. Despite its main focus is on temporal data, the mindset of abstracting out data representation to construct visualisation is applicable to our spatio-temporal data manipulation and visualisation.

@@

This section will first introduce the core functions in the cubble class: \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()} from subsection \ref{create} to \ref{unfold}. The next subsection addresses why bother a new class before subsection \ref{compact} finishing on the compatibility of cubble with existing packages in spatial and temporal analysis.

Each core function is introduced with a short example accompanied and the data \texttt{climate\_flat} is used throughout this section. \texttt{climate\_flat} contains five weather stations in Australia with spatial information of each station: station id, latitude, longitude, elevation, station name and World Meteorology Organisation ID and also daily temporal information of date, maximum and minimum temperature and precipitation for 2020. Below prints out the first five rows of \texttt{climate\_flat}:

@@

\hypertarget{create}{%
\subsection{Create a cubble in the nested form}\label{create}}

As described in the introduction, spatio-temporal data can come in various formats and shapes. A cubble can be created from various raw data, including tibble and its variates multiple tables (an example provided in section \ref{covid}), and netcdf (detailed in section \ref{netcdf}). The function \code{as_cubble()} is used to create a cubble with three additional components: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order of longitude and latitude. Argument \code{key} and \code{index} follow the wording in \pkg{tsibble} to define temporal order and multiple series while \code{coords} is a space specific argument to locate each site in two columns. The code below creates a cubble out of \code{climate_flat} with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

@@

The cubble header provides some information about this data. \code{id} is the variable name to identify each location and there are five unique locations. The bounding box is \code{[115.97, -32.94, 133.55, -12.42]} and the names of variable nested in the \code{ts} column, along with its type, are \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

The created cubble is built from a \code{rowwise_df} class where each row forms a group and all the temporal variables are nested in a list column, hence it is also called the nested cubble. The rowwise structure makes it simpler to calculate on the list using \code{mutate()} and avoid the \code{map()} when working with list column. For example calculating the number of raining day can be done by:

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts       rain_day
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>      <int>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble>      104
2 ASN00010311 -31.9  117. 179   york            94623 <tibble>       89
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble>       90
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble>      106
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble>       63
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{switch-from-the-nested-cubble-into-the-long-form}{%
\subsection{Switch from the nested cubble into the long form}\label{switch-from-the-nested-cubble-into-the-long-form}}

The nested format is convenient for those operations whose output does not contain a time dimension. For those outputs that are cross-identified by the spatial and temporal identifier, a long cubble needs to be used. The function \code{face_temporal()} switches the cubble from the nested form to the long form and the first row in Figure \ref{fig:face} sketches the function in a diagram. This code shows how to switch the cubble just created into its long form:

@@

Notice that the third line in the header now shows the name and type of spatial variables: \code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]} and these variables are now stored as a \code{spatial} attribute of the data:

\begin{CodeChunk}
\begin{CodeInput}
R> attr(cubble_long, "spatial")
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 5 x 6
  id            lat  long  elev name           wmo_id
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610
2 ASN00010311 -31.9  117. 179   york            94623
3 ASN00010614 -32.9  117. 338   narrogin        94627
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120
5 ASN00015131 -17.6  134. 220   elliott         94236
\end{CodeOutput}
\end{CodeChunk}

The long form is built form class \code{groupped_df} where all the observations from the same sites forms a group.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.008}

@@

\caption{An illustration of function \code{face\_temporal} and \code{face\_spatial} in cubble. In the first row, \code{face\_temporal} switches a cubble from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \code{face\_spatial} switches a cubble back to the nested form from te long form and shift focus back to the spatial aspect.}\label{fig:face}
\end{figure}
\end{CodeChunk}

\hypertarget{switch-a-long-cubble-back-to-the-nested-form}{%
\subsection{Switch a long cubble back to the nested form}\label{switch-a-long-cubble-back-to-the-nested-form}}

Manipulation on the spatial and temporal dimension can be an iterative process and analysts may need to go back and forth between the nested and long cubble. \code{face_spatial()}, the inverse of \code{face_temporal()}, switches the long cubble back to the nested form, as sketched in the second row in Figure \ref{fig:face}:

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_back <- cubble_long |> face_spatial()
R> cubble_back
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{unfold}{%
\subsection{Flatten spatial variables into the long cubble}\label{unfold}}

Sometimes, analysts may need to apply some variable transformation that involves both the spatial and temporal variable. An example of this is the transformation of temporal variables into the spatial dimension in glyph maps, which will be elaborated in section \ref{st_transformation}. This type of operations can be seen as flattening, or \emph{unfolding}, the cube into a 2D data frame. Here the function \code{unfold()} moves the spatial variables \code{long} and \code{lat} into the long cubble:

@@

This function should generally be used in the last step of the analysis since it is a temporary operation, meaning these added spatial variables are not stored in the long form and will disappear if switched to the nested form and then switched back:

@@

\subsection[Why not just use]{Why not just use \code{dplyr::nest\_by()}}

It may naturally raise the question that why bother a new data structure since the nested form can already be created by the \pkg{dplyr} function \code{nest_by()}? While \code{nest_by()} and \code{unnest()} can mimic the nested form in cubble, cubble can be seen as an attempt to arrange variables from different observational units into a single object, specifically for spatio-temporal data. While \texttt{nest\_by()} solves this problem on the spatial side, it doesn't solve the temporal side. For temporal data, a combined representation with spatial variables is useful for spatio-temporal operations but its duplication can be a burden to operate on. Also, spatial variables are dropped when summarised and you will need to either specified all of them in the \code{group_by()} before \code{summarise}, or add a joining after \code{summarise()}.

@@

In the nested cubble created, each element in the list-column \code{ts} is of \code{tbl_ts} class and operations available to the tsibble class is still valid under cubble. For example, the code below calculates two features of the maximum temperature:

@@

As a spatial data object, \code{sf} creates a simple feature geometry list-column (\code{sfc}) in the data frame to provide spatial operations on various geometry types (\code{POINT}, \code{LINESTRING}, \code{POLYGON}, \code{MULTIPOLYGON}, etc). These spatial operations are also valuable for spatio-temporal data analysis, but an \code{sf} object \emph{usually} does not contain temporal variables. This means \code{sf} cannot be directly cast into a \code{cubble}, however, \code{cubble} does support \code{sfc} columns in the nested form and spatial operations applied to the \code{sfc} column in \code{sf} can still be applied to the \code{sfc} column in a cubble. The following example shows how to create an \code{sfc} column of \code{POINT} type from latitude and longitude in cubble. Then \code{sf::st_within} is used to add the state \code{MULTIPOLYGON} of each weather station before a coordinate transformation is made.

\begin{CodeChunk}
\begin{CodeInput}
R> library(sf)
R> # create a cubble
R> cb <- climate_flat |>
+   cubble::as_cubble(key = id, index = date, coords = c(long, lat))
R> 
R> aus <- ozmaps::abs_ste
R> 
R> dt <- cb |>
+   mutate(
+     # create `sfc` column based on long and lat
+     ll = st_sfc(
+       purrr::map2(long, lat, ~st_point(c(.x, .y))),
+       crs = st_crs(aus)),
+ 
+     # append state multi-polygon based on the `sfc` created
+     state = aus$geometry[st_within(ll, aus, sparse = FALSE)],
+ 
+     # adopt a different projection: lambert conformal conic (EPSG:3112)
+     state = st_transform(state, crs = "EPSG:3112")
+     )
R> 
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name   wmo_id ts                        ll
  <chr>       <dbl> <dbl> <dbl> <chr>   <dbl> <list>           <POINT [°]>
1 ASN00009021 -31.9  116.  15.4 perth~  94610 <tibble> (115.9764 -31.9275)
2 ASN00010311 -31.9  117. 179   york    94623 <tibble>  (116.765 -31.8997)
3 ASN00010614 -32.9  117. 338   narro~  94627 <tibble> (117.1797 -32.9342)
4 ASN00014015 -12.4  131.  30.4 darwi~  94120 <tibble> (130.8925 -12.4239)
5 ASN00015131 -17.6  134. 220   ellio~  94236 <tibble> (133.5407 -17.5521)
# ... with 1 more variable: state <MULTIPOLYGON [m]>
\end{CodeOutput}
\end{CodeChunk}

An \pkg{s2} \code{lnglat} vector can similarly be created as an \code{sfc} in cubble before using any \code{s2}-prefixed function:

\begin{CodeChunk}
\begin{CodeInput}
R> library(s2)
R> # Western Australia map
R> wa <- ozmaps::abs_ste |> filter(NAME == "Western Australia")
R> 
R> # mutate a `s2_lnglat` vector on `cb` created in the last chunk
R> cb |>
+   mutate(ll = s2_lnglat(long, lat)) |>
+   filter(s2_within(ll, wa))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [3]: nested form
# bbox:     [115.97, -32.94, 117.18, -31.89]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name          wmo_id ts                 ll      
  <chr>       <dbl> <dbl> <dbl> <chr>          <dbl> <list>             <s2_lng>
1 ASN00009021 -31.9  116.  15.4 perth airport  94610 <tibble [366 x 4]> (115.97~
2 ASN00010311 -31.9  117. 179   york           94623 <tibble [366 x 4]> (116.76~
3 ASN00010614 -32.9  117. 338   narrogin       94627 <tibble [366 x 4]> (117.17~
\end{CodeOutput}
\end{CodeChunk}

@@

NetCDF data has two main components: \emph{dimension} for defining the spatio-temporal grid (longitude, latitude, and time) and \emph{variable} that populates the defined grid. Attributes are usually associated with dimension and variable in the NetCDF format data and a \href{http://cfconventions.org/}{metadata convention for climate and forecast} has been designed to standardise the format of the attributes. A few packages in R exists for manipulating NetCDF data and this includes a high-level R interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls C-interface: \pkg{RNetCDF} \citep{rnetcdf, michna2013rnetcdf}, and a tidyverse implementation: \pkg{tidync} \citep{tidync}.

@@

The memory limit with NetCDF data in cubble depends on the number of longitude grid points times the number of latitude grid points times the number of time grid points times the number of variables. Cubble can handle slightly more than 300 x 300 (longitude x longitude) grid points for 3 variables in one year. The spatial grid can be reduced to trade for longer time periods and more variables. A 300 by 300 spatial grid can be a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or a global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution. Subsetting longitude and latitude grids is available through \code{long_range} and \code{lat_range} if the NetCDF file has finer resolution than needed.

\begin{CodeChunk}
\begin{CodeInput}
R> # Assume my_ncdf has bounding box of [-180, -90, 180, -90]
R> # at 0.25 degree resolution and subset it to have
R> # 1 degree resolution:
R> dt <- as_cubble(my_ncdf, vars = c("q", "z"),
+                 long_range = seq(-180, 180, 1),
+                 lat_range = seq(-90, 90, 1))
\end{CodeInput}
\end{CodeChunk}

@@

Spatial locations can have grouping structure either inherent to the data or formed by clustering. Rather than analysing variables in the site level, summarised variables in the cluster level can give a crisper picture of local areas. In cubble, \code{switch_key()} can be used to create a new level of grouping of spatial locations by specifying a clustering variable. Figure \ref{fig:illu-hier} illustrates the relationship of cubbles at station and cluster level, in both the long and nested form. By specifying \code{cluster_nested <- station_nested \%>\% switch_key(key = cluster)}, the cubble re-defines the cubble key from \code{id} in \code{station_nested} to \code{cluster} in \code{cluster_nested}. All the spatial variables variant to \code{cluster} are now nested into a \code{.val} column and cluster level variables can be computed in the same fashion as station level variables in \code{station_nested}.

@@

{\centering \includegraphics[width=1\linewidth,height=0.4\textheight]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.003}

@@

\hypertarget{data-fusion-and-matching}{%
\subsection{Data fusion and matching}\label{data-fusion-and-matching}}

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously.
In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.

The SpatioTemporal CRAN task view \citep{ctvspatiotemporal} gathers information about R packages designed for spatio-temporal data and it has a section on \emph{Representing data} that lists existing spatio-temporal data representations used in \proglang{R}. Among them, \citet{spacetime} summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} \citep{spacetime} implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The \pkg{stars} \citep{stars} package has a new implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with \pkg{sf} \citep{sf}, a package commonly used for wrangling spatial data, and the \pkg{tidyverse} \citep{tidyverse} suite for general data wrangling and visualisation in \proglang{R}.

Still, the data representation for spatio-temporal data can be further extended and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the layouts above, and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) areal data recording the shape of a collection of areas of interest; 2) geostatistical data storing the longitude and latitude coordinates of locations, typically also with other metadata related to the location, and; 3) temporal data of each location across time.

The other reason is about tidy data concepts \citep{tidydata} and how they should be applied to spatio-temporal data. According to the tidy data principles, data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data form given the downstream software such as \pkg{dplyr} \citep{dplyr} and \pkg{ggplot2} \citep{ggplot2} for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for hourly, daily or sub-daily periods over years, which are extensively collected and handled, for example in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make data wrangling, visualizing and analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in \citet{spacetime}, \pkg{cubble} can be applied to full grid, sparse grid, or irregular, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=cubble}.

The rest of the paper is organized as follows: Section \ref{cube} introduces the proposed cube structure as a way to conceptualise multivariate spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations, including data with hierarchical structure, data matching and how cubble fits with existing static and interactive visualisation tools. Moreover we also illustrate how \pkg{cubble} deals with spatio-temporal data transformations. Section \ref{examples} uses Australian weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} discuss the paper contributions and future directions.

@@

Spatio-temporal data can be conceptualised using a cubical data model with three axes which typically are, time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation purposes: \citet{lu_multidimensional_2018} shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; \citet{bach_review_2014} reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data'', but refers to ``an abstract 2D substrate that is used to visualize data at a specific time''. Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisations, still applies to our spatio-temporal data manipulation and visualisation approaches.

The most common data cube using the three axes (time, latitude, longitude) could be considered as taking snapshots of the space and stacking them across time. Here we define the spatio-temporal cube slightly differently by considering the three axes to be, time, site and variable/s. That is illustrated in the leftmost cube in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the conceptual framework behind \pkg{cubble}. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.

While the data cube model is conceptually convenient for spatio-temporal data, a 3D data array is not sufficiently rich for data wrangling, for several reasons. Although arrays can be efficient for the computation on numerical values, spatio-temporal data typically includes various types of variables. For example, character strings and specific datetime classes are common. In addition, it will be generally useful to be able to create new variables which is trickier to manage in an array. Thus for convenient wrangling, we have opted for a list object as the \texttt{cubble} structure.

{\centering \includegraphics[width=1\linewidth]{../figures/diagram-keynotes/diagram-keynotes.007}

\caption[An illustration of the conceptual spatio-temporal cube with different slicing on time, site, and variable]{An illustration of the conceptual spatio-temporal cube with different slicing on time, site, and variable. For each axis, the slicing can be on a single value or a set of values.}\label{fig:cube-def}
\end{figure}
\end{CodeChunk}

@@

Here we introduce the core functionalities of cubble. The main functions, \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()} are described in Section \ref{create} - \ref{unfold}. Section \ref{tidyverse} explains the need to create a new class. Section \ref{compact} will then discusses how cubble relates to existing packages for spatial and temporal data analysis.

The dataset \texttt{climate\_flat} is used to illustrate \pkg{cubble} functionality. This is a subset from the ORIGIN and contains the spatial variables, station id, latitude, longitude, elevation, station name, World Meteorology Organisation ID, in additional to daily temporal information, maximum and minimum temperature values and precipitation records for 2020. The first five rows of \texttt{climate\_flat} are shown below:

@@

\hypertarget{create}{%
\subsection{Create a cubble}\label{create}}

A cubble can be created from various common spatio-temporal data formats, including basic R objects like tibble, tsibble, sf. Section \ref{covid} describes converting multiple tables into a cubble object and Section \ref{netcdf} illustrate how to convert a netCDF object.

The function \code{as_cubble()} is used to create a cubble with three arguments: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order (longitude, latitude). The arguments \code{key} and \code{index} follow the wording in \pkg{tsibble} to describe the temporal order and multiple series while \code{coords} specifies the spatial location of each site. The code below creates a cubble out of \code{climate_flat} (a single tibble) with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

@@

Printing a cubble provides some information about the data. \code{id} is the variable name to identify each location and there are five unique locations. The bounding box is \code{[115.97, -32.94, 133.55, -12.42]} and provides information about the coordinates in the data set. The third row shows the name and type of all variables nested in the \code{ts} column. In this example, it includes \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

A cubble object is an extension of a \code{rowwise_df} object where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested cubble. The rowwise structure makes it simpler to operate on the list using the \code{mutate()} syntax, which is simpler than the \code{purr::map()} when working with a list column. For example, calculating the number of rainy days can be done by:

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_nested |>
+   mutate(rain_day = sum(ts$prcp != 0))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts       rain_day
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>      <int>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble>      104
2 ASN00010311 -31.9  117. 179   york            94623 <tibble>       89
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble>       90
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble>      106
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble>       63
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{change-focus-by-facing-the-time-variables}{%
\subsection{Change focus by facing the time-variables}\label{change-focus-by-facing-the-time-variables}}

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). The \pkg{cubble} class also provides a long form, which expands the \code{ts} column and temporarily ``hides'' the spatial variables. The function \code{face_temporal()} is used to switch from the nested cubble into the long one. The first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the cubble just created into its long form:

@@

The first line in printed cubble now shows the cubble in its long form and the third line has been changed to display the name and type of spatial variables: \code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}. Unlike the nested form, the long cubble is built from a \code{grouped_df} object where all the observations from the same sites form a group.

@@

{\centering \includegraphics[width=1\linewidth]{../figures/diagram-keynotes/diagram-keynotes.008}

@@

\caption{An illustration of function \code{face\_temporal} and \code{face\_spatial} in cubble. In the first row, \code{face\_temporal} switches a cubble from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \code{face\_spatial} switches a cubble back to the nested form from the long form and shifts focus back to the spatial aspect.}\label{fig:face}
\end{figure}
\end{CodeChunk}

\hypertarget{change-focus-back-to-the-site-variable-face}{%
\subsection{Change focus back to the site-variable face}\label{change-focus-back-to-the-site-variable-face}}

Wrangling spatio-temporal data can be seen as an iterative process in the spatial and temporal dimensions. Switching the focus back to the site-variable face can be accomplished by the function \code{face_spatial()}, which is the inverse of \code{face_temporal()}. The second row of Figure \ref{fig:face} illustrates the function, which is used as follows:

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_back <- cubble_long |> face_spatial()
R> cubble_back
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{CodeOutput}
\begin{CodeInput}
R> identical(cubble_nested, cubble_back)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{unfold}{%
\subsection{Unfold spatial variables into the long cubble}\label{unfold}}

Sometimes, analysts may need to apply some variable transformation that involves both the spatial and temporal variables. An example of this is the transformation of temporal variables into the spatial dimension in glyph maps {[}ref{]}. (Glyph maps will be explained in Section \ref{st_transformation}.) This type of operation can be seen as flattening, or \emph{unfolding}, the cube into a 2D data frame. Here the function \code{unfold()} moves the spatial variables \code{long} and \code{lat} into the long cubble:

@@

This function should generally be used in the last step of the analysis since it is a temporary operation, meaning that these added spatial variables will disappear, if switched to the nested form and then switched back:

@@

\hypertarget{tidyverse}{%
\subsection{Why not just use the existing tidyverse functions}\label{tidyverse}}

Some readers may question why a new data structure is needed rather than using \code{dplyr::nest_by()} on the combined data to directly create a list-column. The reason is that cubble is specifically designed to utilize the spatio-temporal structure when arranging variables from different observational units into a single object. Moreover, it enables easy pivoting between purely spatial, purely temporal, or unfolded into a combined form.

@@

When a nested cubble is created, each element in the list-column \code{ts} is of \code{tbl_ts} class and operations available to the tsibble class are still valid within the cubble. For example, the code below calculates two time series features (mean and variance) of the maximum temperature variable:

@@

As a spatial data object, \code{sf} creates a feature geometry list-column (\code{sfc}) in the data frame to provide spatial operations on various geometry types such as (\code{POINT}, \code{LINESTRING}, \code{POLYGON}, \code{MULTIPOLYGON}, etc). When creating a cubble a \code{sf} object can be supplied as the spatial table. Once that is included in a cubble, methods for the \code{sfc} class can be applied in the nested form inside the cubble object. An example of this is shown in \ref{covid}, which also handles the case when the site identifiers in the two tables do not match exactly. A spatial data object with an \code{s2} vector can also be the input for the spatial table in \pkg{cubble}.

@@

NetCDF data is another format that is commonly used for storing spatio-temporal data. It has two main components: \emph{dimension} for defining the spatio-temporal grid (longitude, latitude, and time) and \emph{variable} that populates the defined grid. Attributes are usually associated with dimensions and variables in the NetCDF format data and a \href{http://cfconventions.org/}{metadata convention for climate and forecast} has been designed to standardise the format of the attributes. A few packages in R exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls a C-interface: \pkg{RNetCDF} \citep{rnetcdf, michna2013rnetcdf}, and a tidyverse implementation: \pkg{tidync} \citep{tidync}.

@@

The memory limit with NetCDF data in cubble depends on the number of longitude grid points \(\times\) the number of latitude grid points \(\times\) the number of time grid points \(\times\) the number of variables. Cubble can handle slightly more than 300 \(\times\) 300 (longitude \(\times\) longitude) grid points for three daily variables in one year. A 300 by 300 spatial grid can be a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or a global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution. The spatial grid can also be subsetted to trade for longer time periods and more variables through \code{long_range} and \code{lat_range} if the NetCDF file has finer resolution than needed:

\begin{CodeChunk}
\begin{CodeInput}
R> # Assume my_ncdf has a bounding box of [-180, -90, 180, -90]
R> # at 0.25 degree resolution and subset it to have
R> # 1 degree resolution:
R> dt <- as_cubble(my_ncdf, vars = c("q", "z"),
+                 long_range = seq(-180, 180, 1),
+                 lat_range = seq(-90, 90, 1))
\end{CodeInput}
\end{CodeChunk}

@@

Spatial locations can have grouping structures either inherent to the data or obtained by clustering. In those cases, rather than analysing variables at the site level, it might be of interest to summarised variables at a cluster or group level, as that can give a crisper picture of local areas. In cubble, \code{switch_key()} can be used to create a new grouping level of spatial locations by specifying a clustering variable. Figure \ref{fig:illu-hier} illustrates the relationship of cubbles at station and cluster level, in both the long and nested forms. By specifying \code{cluster_nested <- station_nested \%>\% switch_key(key = cluster)}, the cubble redefines the cubble key from the \code{id} column in \code{station_nested} to the \code{cluster} column in \code{cluster_nested}. All the spatial variables variant to the \code{cluster} column are now nested into a \code{.val} column and cluster level variables can be computed in the same fashion as station level variables in \code{station_nested}.

@@

{\centering \includegraphics[width=1\linewidth,height=0.4\textheight]{../figures/diagram-keynotes/diagram-keynotes.003}

@@

\hypertarget{matching}{%
\subsection{Data fusion and matching}\label{matching}}
